

# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
  library(tidyverse)   # Data wrangling
  library(here)        # File management
  library(readxl)      # Read excel files
  library(sf)          # Spatial processing
  library(tmap)        # Maps
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())
wgs_crs = 4326
ca_crs = 3310

tmap_mode(mode = "view")
```

# Import data

Cleaned & deduplicated network data
```{r}
network = read_excel("data/Source/Network data/Final data 10.22.25_latlonadded.xlsx", na = "NA")
```

# Cleaning and processing

- Remove columns we definitely don't need
- Rename columns and reorder remaining columns
- Convert any columns with 0/1 to No/Yes, or in the case of gw, combine both columns into one categorical column
- Add new column for system size based on population served (thresholds provided by Kristin)
- Clean up errors in source type column
- Standardise all source lat/longs (there are currently different values for the same source)

```{r}
network_clean = network%>%
  rename(pop_served = Population_served_count, source_name = Source)%>%
  rename_with(tolower)%>%
  
  mutate(
    gw_access = case_when(
      gw_access_no_cc == 0 ~ "No Groundwater",
      gw_compliant_w_dw_standards_no_cc == 1 ~ "Groundwater access with 1 or more well meeting MCL requirements",
      gw_compliant_w_dw_standards_no_cc == 0 ~ "Groundwater access with no wells meeting MCL requirements"),
    system_size = case_when(
      pop_served == 0 ~ NA,
      pop_served > 100000 ~ "Very Large",
      pop_served > 10000 ~ "Large",
      pop_served > 3300 ~ "Medium",
      pop_served > 500 ~ "Small",
      pop_served <= 500 ~ "Very Small"),
    system_lat = as.numeric(system_lat),
    swp_loop = ifelse(swp_loop == 0, "No", "Yes"),
    cvp_loop = ifelse(cvp_loop == 0, "No", "Yes"),
    source_type = case_when(
      source_name == "Kern River" ~ "River",
      source_name == "Lake Havasu" ~ "Lake",
      source_name == "Lake Spaulding" ~ "Lake",
      T ~ source_type))%>%
  
  group_by(source_name)%>%
  mutate(num_systems_served = n(),
         source_lat = first(na.omit(source_lat)),
         source_lon = first(na.omit(source_lon)))%>%
  ungroup()%>%
  
  group_by(pwsid)%>%
  mutate(num_sw_sources = n())%>%
  ungroup()%>%

  select(-c(is_wholesaler, source_number, source_summary, source_id, combined_source_notes, combined_calsim_notes, swp:colorado_incompleteandnotedited, x2020_ccr_percent:x2023_percent, summer_2025_cleaning, calsim_demandunit:calsim_notes_sept2025, colorado_loop, connection_id, gw_access_no_cc:gw_compliant_w_dw_standards_no_cc))%>%
  
  select(1:2,5:6,18,7,17,20,3:4, 8:9,19,10:11, 12:16)%>%
  rename(swp = swp_loop, cvp = cvp_loop)
```

# Split into 3 datasets

## Systems

Keep one row per system
```{r}
systems_sf = network_clean%>%
  select(system_name:system_lon)%>%
  filter(!is.na(system_lat))%>%
  unique()%>%
  st_as_sf(., coords = c("system_lon", "system_lat"), crs = wgs_crs)%>%
  st_transform(ca_crs)

tm_shape(systems_sf%>%filter(pwsid == "CA3710707"))+
  tm_dots()
```


## Sources

Keep one row per source, only selecting the first in each group (this is because multiple slightly different lat/longs are recorded for most source, so just taking the first one). 

Exclude any PWS sources since these will be in the systems dataset.

Convert to sf point object.
```{r}
sources_sf = network_clean %>%
  select(source_name, source_type, source_lat, source_lon) %>%
  filter(!is.na(source_lat) & source_type != "PWS") %>%
  unique()%>%
  st_as_sf(coords = c("source_lon", "source_lat"), crs = wgs_crs) %>%
  st_transform(ca_crs)

tm_shape(sources_sf)+
  tm_dots()
```

## Connections

Generate linestring from master connections dataframe based on source/system coordinates
```{r}
connections_sf = network_clean %>%
  filter(!is.na(source_lat) & !is.na(system_lat)) %>%
  rowwise() %>%
  mutate(
    geometry = list(st_linestring(
      matrix(
        c(source_lon, source_lat, system_lon, system_lat), 
        nrow = 2, 
        byrow = TRUE)))) %>%
  st_as_sf(crs = wgs_crs) %>%
  st_transform(ca_crs)

tm_shape(connections_sf)+
  tm_lines()+
  tm_shape(systems_sf)+
  tm_dots(fill = "skyblue3", size = 0.25)+
  tm_shape(sources_sf)+
  tm_dots(fill = "coral2", size = 0.25)
```

# Export
```{r}
st_write(connections_sf, "Data/Outputs/Shapefiles for dashboard/20251023_connections.shp", delete_dsn = T)
st_write(systems_sf, "Data/Outputs/Shapefiles for dashboard/20251023_systems.shp", delete_dsn = T)
st_write(sources_sf, "Data/Outputs/Shapefiles for dashboard/20251023_sources.shp", delete_dsn = T)
```





