
# Overview

The team has constructed a network dataset of California public water systems and the various sources they get their surface water from. 

In this script, we want add two groundwater-related variables to the dataset (for now in a separate table that we can join by PWSID). One being a simple binary just whether the PWS has groundwater or not, and the second being some indicator of if their groundwater meets drinking water standards.


# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
  library(tidyverse)   # Data wrangling
  library(here)        # File management
  library(readxl)      # Read excel files
  library(janitor)     # Clean up dataframe column names
  library(lubridate)   # Extract years from dates
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())
```


# Import data

## Network data
Draft version of a network dataset of public water systems compiled by the team. 

```{r}
network_raw = read_excel("Data/Source/Draft_networkdata_Summer2025_Aug25.xlsx")
```

Note that the data is by source, so PWSIDs are repeated for as many water sources as the system has. 
Since we want the final output to just be a dataframe of the 2 groundwater variables by PWSID (which we can join back into the master dataframe by PWSIDlater once it's finalised), let's take a vector of the unique PWSIDs
```{r}
nrow(network_raw)
n_distinct(network_raw$PWSID)

network_pwsids = data.frame(pwsid = unique(network_raw$PWSID))

network_pwsids%>%
  filter(!grepl("CA", pwsid))
```
Although the dataset has 1117 rows, there are 861 unique PWSIDs, which we've now saved into a vector without duplicates. 

Note that there are 4 PWSs that aren't regulated any therefore do not have a 'normal' PWSID. Flagging as we will be assigning these values of NA for the groundwater variables.

## SAFER clearinghouse data
This is an internal state water board dataset they shared with us for this project. All documentation is located in [this Google Drive folder](https://drive.google.com/drive/folders/1gxMrGeaOA_xtGb1zf97Z79EFEwKcIpCG).

Unfortunately there is no data dictionary, but [Jenny has worked on developing one](https://docs.google.com/spreadsheets/d/1ulStajU7D1eFu2m3V0fTDtcm3PGNlneX/edit?gid=631545570#gid=631545570).

```{r}
# read.csv("Data/Source/SAFER clearinghouse/active_cws_04_2025.csv")
# clearinghouse_raw = read_excel("Data/Source/SAFER clearinghouse/20250424_SAFER_CLEARINGHOUSE_EXPORT.xlsx", sheet = "SourceReporting")

#save(clearinghouse_raw, file = "Data/Source/SAFER clearinghouse/safer_clearinghouse.RData")
load("Data/Source/SAFER clearinghouse/safer_clearinghouse.RData")

clearinghouse_raw
```

We want to filter for facilities with likely groundwater access. To do so, we will use the following filters:
- Only active facilities
- Exclude facilities serving as an emergency water source
- Filter out any sources that are explicitly surface water, in both the clearinghouse and SDWIS water type columns
- Filter out any springs, as these are considered surface water in our network dataset

```{r}
clearinghouse_filtered = clearinghouse_raw%>%
  filter(facility_activity_status == "Active" & 
           sdwis_water_type != "Surface Water" & 
           !clearinghouse_water_type %in% c("Surface Water", "Spring Water") &
           facility_availability != "Emergency" &
           facility_type != "Spring")%>%
  mutate(index_facility_id = str_remove_all(index_facility_id, " "))%>%
  select(water_system_id, facility_name, facility_id, index_facility_id, facility_availability, facility_activity_status:clearinghouse_water_type)
```

Diagnostics to send to Kristin for review
```{r}
check = clearinghouse_filtered%>%
  filter(grepl("Spring", facility_name, ignore.case = T))%>%
  arrange(water_system_id, facility_id)%>%
  unique()

check

write.csv(check, "Data/Outputs/20250902_facilities_to_check.csv", row.names = F)
```

## EDT

Drinking water quality data compiled by the State Water Resources Control Board and downloaded from [this page](https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/EDTlibrary.html). Data dictionary is available [here](https://www.waterboards.ca.gov/drinking_water/certlic/drinkingwater/documents/edtlibrary/data_dictionary.pdf).

Similarly, with the EDT data we want to filter for MCL violations, in addition to the groundwater-related filters in the SAFER clearinghouse data. We will only consider violations in the most recent dataset, containing data from Jan 1, 2023 to Present (last updated August 19, 2025)

```{r}
#edt = read_delim("Data/Source/EDT Library/SDWIS4/SDWIS4.tab", delim = "\t", escape_double = FALSE, trim_ws = TRUE)%>%
  #clean_names()
#save(edt, file = "Data/Source/EDT Library/edt4.Rdata")
load("Data/Source/EDT Library/edt4.Rdata")
```

Filter for analyses with results in which the contaminant met or exceeded MCL levels

```{r}
edt_mcl = edt%>%
  filter(result >= mcl)%>%
  mutate(ps_code_clean = str_remove(ps_code, "_[^_]*$"), mcl_violation = 1)%>%
  select(ps_code_clean, mcl_violation)
```



# Combining datasets

Join clearinghouse data with MCL violations and aggregate by PWSID. 
```{r}
clearinghouse_w_mcl_violations = left_join(clearinghouse_filtered, edt_mcl, by = c("index_facility_id" = "ps_code_clean"))%>%
  mutate(gw_access_facility = 1,
         facility_meets_dw_standards = case_when(
           mcl_violation == 1 ~ 0,
           is.na(mcl_violation) ~ 1))%>%
  group_by(water_system_id)%>%
  reframe(gw_access = max(gw_access_facility), 
          gw_compliant_with_dw_standards = max(facility_meets_dw_standards))

clearinghouse_w_mcl_violations
```
Now join with our original list of PWSIDs and add the new dummy columns, based on:
 - Whether or not PWS has at least 1 facility with groundwater access
 - Whether or not PWS has at least 1 groundwater facility compliant with drinking water standards (which we define as a facility without MCL violations)
```{r}
network_pwsids_w_gw_variables = network_pwsids%>%
  left_join(., clearinghouse_w_mcl_violations, by = c("pwsid" = "water_system_id"))%>%
  mutate(gw_access = case_when(
    !grepl("CA", pwsid) ~ NA,
    is.na(gw_access) ~ 0,
    T ~ gw_access))
```

Inspect to make sure everything looks okay
```{r}
table(network_pwsids_w_gw_variables$gw_access, useNA = "always")
table(network_pwsids_w_gw_variables$gw_compliant_with_dw_standards, useNA = "always")

table(network_pwsids_w_gw_variables$gw_access, 
      network_pwsids_w_gw_variables$gw_compliant_with_dw_standards, 
      useNA = "always",
      dnn = c("GW Access", "GW Compliant with DW Standards"))
```
Export
```{r}
write.csv(network_pwsids_w_gw_variables, "Data/Outputs/20250902_network_pwsids_w_gw_variables.csv", row.names = F)
```





