

# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
  library(tidyverse)   # Data wrangling
  library(here)        # File management
  library(readxl)      # Read excel files
  library(sf)          # Spatial processing
  library(tmap)        # Maps
  library(glue)        # Text processing
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())
wgs_crs = 4326
ca_crs = 3310

tmap_mode(mode = "view")

set.seed(124)
```

# Import data

Cleaned & deduplicated network data
```{r}
network = read.csv("Data/Outputs/20260316_network_joined.csv")
```


# Cleaning and processing

## Initial cleaning

- Remove columns we definitely don't need
- Convert any columns with 0/1 to No/Yes, or in the case of gw, combine both columns into one categorical column
- Add new column for system size based on population served (thresholds provided by Kristin)
```{r}
network_clean = network%>%
  
  mutate(
    purchased = ifelse(purchased == 0, "No", "Yes"),
    source_type = ifelse(source_type == "Lake", "Lake/Reservoir", source_type),
    gw_access_text = case_when(
      gw_access == 0 ~ "No groundwater",
      gw_compliant_w_dw_standards == 1 ~ "Groundwater access with 1 or more well meeting MCL requirements",
      gw_compliant_w_dw_standards == 0 ~ "Groundwater access with no wells meeting MCL requirements",
      is.na(gw_access) ~ "No data"),
    gw_access_filter = ifelse(
      gw_access == 0 | gw_compliant_w_dw_standards == 0, "No drinkable groundwater", NA),
    system_size_filter = case_when(
      is.na(pop_served) ~ "6. NA (Wholesale Provider)",
      pop_served > 100000 ~ "5. Very Large (>100,000)",
      pop_served > 10000 ~ "4. Large (10,000-100,000)",
      pop_served > 3300 ~ "3. Medium (3,300-10,000)",
      pop_served > 500 ~ "2. Small (500-3,300)",
      pop_served <= 500 ~ "1. Very Small (<500)"),
    system_size = str_remove(system_size_filter, "^\\d+\\.\\s"),
    swp_loop = ifelse(swp_loop == 0, "No", "Yes"),
    cvp_loop = ifelse(cvp_loop == 0, "No", "Yes"))%>%

  select(-c(is_wholesaler, source_summary, combined_source_notes, combined_calsim_notes, x2020_ccr_percent:x2023_percent, summer_2025_cleaning, calsim_demandunit:calsim_notes_sept2025, colorado_loop, gw_access:gw_compliant_w_dw_standards, swp_direct:colorado_direct, colorado_text, connection_id, source_id, state_classification))%>%
  
  rename(swp = swp_loop, cvp = cvp_loop, gw_access = gw_access_text)

rm(network)
```

Add dummy variables for if a source has any SWP connections, CVP connections, etc.
```{r}
network_clean = network_clean%>%
  group_by(source_name)%>%
  mutate(any_swp_source = ifelse(any(swp == "Yes"), "Yes", "No"),
         any_cvp_source = ifelse(any(cvp == "Yes"), "Yes", "No"),
         any_purchased_source = ifelse(any(purchased == "Yes"), "Yes", "No"),
         any_gw_source = case_when(
           any(gw_access_filter == "No drinkable groundwater") ~ "No drinkable groundwater"))%>%
  ungroup()%>%
  group_by(pwsid) %>%
  mutate(
    any_swp_system = ifelse(any(swp == "Yes"), "Yes", "No"),
    any_cvp_system = ifelse(any(cvp == "Yes"), "Yes", "No"),
    any_purchased_system = ifelse(any(purchased == "Yes"), "Yes", "No")) %>%
  ungroup()
```


```{r}
purchased_sources = network_clean %>%
  filter(purchased == "Yes") %>%
  pull(source_name) %>%
  unique()

network_clean = network_clean %>%
  group_by(pwsid) %>%
  mutate(
    any_purchased_system = ifelse(pwsid %in% purchased_sources, "Yes", any_purchased_system))%>%
  ungroup()
```


## Lat/long cleanup

Standardise all source lat/longs (there are currently different values for the same source)
```{r}
network_w_lat_longs = network_clean%>%

  group_by(source_name)%>%
  mutate(source_lat = first(na.omit(source_lat)),
         source_lon = first(na.omit(source_lon)))%>%
  ungroup()%>%
  
  mutate(
    source_lat = case_when(
      source_name == "Lake Havasu" ~ 34.29384,
      source_name == "California Aqueduct" ~ 35.712,
      source_name == "Kern River" ~ 35.75573,
      source_name == "Lake Shastina" ~ 41.51761924281496,
      source_name == "Lake Spaulding" ~ 39.33604519822352,
      source_name == "Lyons Reservoir" ~ 38.098002039789485,
      source_name == "Friant-Kern Canal" ~ 36.62401,
      source_name == "Canyon Creek" ~ 39.2,
      source_name == "San Joaquin River" ~ 37.95831775634314,
      T ~ source_lat),
    source_lon = case_when(
      source_name == "Lake Havasu" ~ -114.1686,
      source_name == "California Aqueduct" ~ -120.0103,
      source_name == "Kern River" ~ -118.4228,
      source_name == "Lake Shastina" ~ -122.39281380480435,
      source_name == "Lake Spaulding" ~ -120.63027515989987,
      source_name == "Lyons Reservoir" ~ -120.16573438405824,
      source_name == "Friant-Kern Canal" ~ -119.3066,
      source_name == "Canyon Creek" ~ -120.8,
      source_name == "San Joaquin River" ~ -121.35648835414098,
      T ~ source_lon))

rm(network_clean)
```

Double check to make sure we just have 1 lat/long for each system regardless of whether it's receiving or transferring water
```{r, eval = F}
rbind(
  network_w_lat_longs%>%
    select(pwsid, system_lat, system_lon)%>%
    unique()%>%
    rename(lat = system_lat, lon = system_lon),
  network_w_lat_longs%>%
    filter(source_type == "PWS")%>%
    select(source_name, source_lat, source_lon)%>%
    unique()%>%
    rename(pwsid = source_name, lat = source_lat, lon = source_lon))%>%
  unique()%>%
  arrange(pwsid)%>%
  group_by(pwsid)%>%
  reframe(n = n())%>%
  filter(pwsid %in% as.vector(network_w_lat_longs%>%filter(source_type == "PWS")%>%pull(source_name)))%>%
  filter(n > 1)
  
```


# Split into 3 datasets

## Systems

Keep one row per system
```{r}
systems_sf = network_w_lat_longs%>%
  select(system_name, pwsid, system_lat, system_lon, provider_type, county, pop_served, num_sw_sources, gw_access, gw_access_filter, system_size, system_size_filter, any_swp_system, any_cvp_system, any_purchased_system)%>%
  unique()%>%
  st_as_sf(., coords = c("system_lon", "system_lat"), crs = wgs_crs)
```

For any points that lie outside their relevant SABL polygons, choose a random point in the relevant SABL polygon instead, and overwrite network df so that connections also get updated.
```{r}
sabl = st_read("C:/Users/benji/Desktop/COEQWAL/calsim-outcomes/Inputs/SABL/System_Area_Boundary_Layer.shp")%>%
  rename(pwsid = 1, sabl_system_name = 4)%>%
  group_by(pwsid) %>%
  slice_max(BOUNDARY_T == "Water Service Area", n = 1)%>%
  select(pwsid)%>%
  filter(pwsid %in% systems_sf$pwsid)%>%
  mutate(in_sabl = 1)%>%
  st_transform(ca_crs)

outside = systems_sf%>%
  st_transform(3310)%>%
  st_join(., sabl%>%select(-pwsid), join = st_intersects)%>%
  filter(is.na(in_sabl) & pwsid %in% sabl$pwsid)
  
sabl_points = sabl %>%
  filter(pwsid %in% outside$pwsid & !st_is_empty(geometry)) %>%
  st_point_on_surface(.) %>%
  st_transform(wgs_crs)%>%
  {tibble(
    st_coordinates(.) %>% 
      as.data.frame() %>%
      rename(sabl_lon = X, sabl_lat = Y),
    pwsid = .$pwsid)}

network_w_lat_longs = sabl_points%>%
  right_join(network_w_lat_longs)%>%
  mutate(system_lat = ifelse(is.na(sabl_lat), system_lat, sabl_lat),
         system_lon = ifelse(is.na(sabl_lon), system_lon, sabl_lon))%>%
  select(-c(sabl_lat, sabl_lon))

systems_sf = network_w_lat_longs%>%
  select(system_name, pwsid, system_lat, system_lon, provider_type, county, pop_served, num_sw_sources, gw_access, gw_access_filter, system_size, system_size_filter, any_swp_system, any_cvp_system, any_purchased_system)%>%
  unique()%>%
  st_as_sf(., coords = c("system_lon", "system_lat"), crs = wgs_crs)%>%
  mutate(project = case_when(
    any_swp_system == "Yes" & any_cvp_system == "Yes" ~ "Both",
    any_swp_system == "Yes" & any_cvp_system == "No" ~ "SWP",
    any_swp_system == "No" & any_cvp_system == "Yes" ~ "CVP",
    T ~ NA))
```


```{r, eval = F}
tm_shape(systems_sf)+
  tm_dots()
```


## Sources

Keep one row per source

Exclude any PWS sources since these will be in the systems dataset.

Convert to sf point object.
```{r}
sources_sf = network_w_lat_longs%>%
  select(source_name, source_type, num_systems_served, source_lat, source_lon, any_swp_source, any_cvp_source, any_purchased_source, any_gw_source) %>%
  filter(!is.na(source_lat) & !source_type %in% c("PWS", "Non-PWS wholesaler")) %>%
  unique()%>%
  st_as_sf(coords = c("source_lon", "source_lat"), crs = wgs_crs)%>%
  mutate(source_type_consolidated = ifelse(source_type == "Other supplier", "Other Suppliers", "Natural Sources"))

jittered_sources = sources_sf%>%
  st_transform(ca_crs)%>%
  st_jitter(., amount = 800)%>%
  st_transform(wgs_crs)

rm(sources_sf)
```

```{r, eval = F}
tm_shape(rbind(sources_sf%>%mutate(df = "original"), jittered_sources%>%mutate(df = "jittered")))+
  tm_dots(col = "df", palette = c("red", "blue"))
```

## Connections

Put new jittered source coordinates in master dataset

Change source names to the system name (just for PWSID sources)
```{r}
network_w_lat_longs = jittered_sources %>%
  st_coordinates() %>%  
  as.data.frame() %>%
  rename(source_lon_jittered = X, source_lat_jittered = Y)%>%
  mutate(source_name = jittered_sources$source_name)%>%
  right_join(., network_w_lat_longs, by = "source_name")%>%
  mutate(source_lat = ifelse(is.na(source_lat_jittered), source_lat, source_lat_jittered),
         source_lon = ifelse(is.na(source_lon_jittered), source_lon, source_lon_jittered))%>%
  select(-c(source_lat_jittered, source_lon_jittered))%>%
  
  mutate(source_pwsid = ifelse(source_type == "PWS", source_name, NA))%>%
  left_join(., 
            systems_sf%>%select(system_name, pwsid)%>%st_drop_geometry()%>%rename(source_name_for_systems = system_name), 
            by = c("source_pwsid" = "pwsid"))%>%
  mutate(source_name = ifelse(is.na(source_name_for_systems), source_name, glue("{source_name_for_systems} ({source_pwsid})")))%>%
  select(-source_name_for_systems)
```


Generate linestring from master connections dataframe based on source/system coordinates
```{r}
connections_sf = network_w_lat_longs %>%
  mutate(system_lat = as.numeric(system_lat), source_lat = as.numeric(source_lat))%>%
  filter(!is.na(source_lat))%>%
  rowwise() %>%
  mutate(
    geometry = list(st_linestring(
      matrix(
        c(source_lon, source_lat, system_lon, system_lat), 
        nrow = 2, 
        byrow = TRUE)))) %>%
  st_as_sf(crs = wgs_crs)%>%
  select(-c(any_swp_source:source_pwsid))

rm(network_w_lat_longs)
```

```{r, eval = F}
tm_shape(connections_sf)+
  tm_lines()+
  tm_shape(systems_sf)+
  tm_dots(fill = "coral2", size = 0.4)+
  tm_shape(jittered_sources)+
  tm_dots(fill = "skyblue3", size = 0.4)
```


# Export
```{r}
st_write(connections_sf, "Data/Outputs/Shapefiles for dashboard/20260327_connections.shp", delete_dsn = T)

st_write(systems_sf, "Data/Outputs/Shapefiles for dashboard/20260327_systems.shp", delete_dsn = T)
#st_write(systems_sf%>%filter(!is.na(project)), "Data/Outputs/Shapefiles for dashboard/20260325_project_systems.shp", delete_dsn = T)

st_write(jittered_sources, "Data/Outputs/Shapefiles for dashboard/20260327_sources.shp", delete_dsn = T)
```



